An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models

### 2025-10-01 Project_0/manuscript/manuscript_R1.Rmd
saving work on project 0 revisons

State-space models are increasingly used for stock assessment, and evaluations of their statistical reliability and best practices for selecting among process error configurations are needed. We simulated 72 operating models that varied fishing pressure and observation error across process errors in recruitment, survival, selectivity, catchability, and/or natural mortality. We fit estimating models with different assumptions on the process error source and whether median natural mortality or a stock-recruit relationship were estimated. Estimating models without a stock-recruit relationship that assumed the correct process error source and median natural mortality had high convergence rates and low bias. Bias was also low under many incorrect process error assumptions when there was contrast in fishing pressure and low observation error. Marginal AIC most accurately distinguished process errors on recruitment, survival, and selectivity, and other process error sources when variability was greater. Retrospective patterns were generally small but were sizable for recruitment when observation error was high. These results help establish the statistical reliability of state space assessment models and pave the way for the next-generation of fisheries stock assessment.

@@

Subsets of 20 EMs in Table \ref{em_table} were fit to simulate data sets from each of the OM process error categories. For R and R+S OMs, fitted EMs had matching process error assumptions as well as R+Sel, R+M, and R+q assumptions without autocorrelation. Similarly, For other OM process error categories, we fit EMs with matching process error assumptions as well as other process error types without autocorrelation. The maturity at age, weight at age for catch and SSB, and observation error variance of aggregate catch and indices were all assumed known at the true values. However, the variance parameters for the logistic-normal distributions for age composition observations were estimated in the EMs. As such, EMs would either be configured completely correctly for the OM, or there could be mis-specification in assumptions of process error autocorrelation, the type of process error, or the SRR (Beverton-Holt or none).

@@

The first measure of reliability we investigated was frequency of convergence when fitting each EM to the simulated data sets. There are various ways to assess convergence of the fit [e.g., @carvalhoetal21; @kapuretal24], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probablity of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted. We also provide results in the Supplementary Materials for the maximum of the absolute values among all gradients for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects. This provides an indication of how poor the calculated gradients can be, but still presumably converged adequately enough for parameter inferences. We used the Clopper-Pearson exact method for constructing 95\% confidence intervals of the probabilities of convergence [@clopperpearson34;@thulin14].

@@

We estimated the probability of selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given operating model, we compared AIC for EMs that were all configured the same for median natural mortality (known or estimated) and the SRR (Beverton-Holt or none).

We also estimated the probability of correctly selecting between EMs with Beverton-Holt SRR assumed and models without the SRR (null model). We made these comparisons between models that otherwise assumed the same process error structure as the operating model and both of the compared models either estimate median natural mortality or assume it is known. Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimation of SRR parameters [@magnussonhilborn07; @connetal10]. Our preliminary inspections of the proportions of simulations where the correct recruitment model was chosen for a given set of OM factors (including contrast in fishing pressure) indicated generally poor performance of AIC. Therefore, we fit logistic regression models to the indicator of Beverton-Holt models having lower AIC as a function of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) since simulations with realized SSB producing low and high recruitments would have larger variation in realized SSB.

All model selection results condition only on whether all of the compared estimating models completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization could correctly determine an inappropriate process error assumption by estimating variance parameters at the lower bound of zero. Such an optimization could indicate poor convergence but the likelihood would be equivalent to that without the mis-specified random effects and the AIC would be appropriately higher because more (variance) parameters were estimated. All other measures of reliability described below (bias and Mohn's $\rho$) use these same criteria for inclusion of EM fits in the summarized results.

@@

For a given model attribute we calculated the relative error 
\begin{equation}\label{relerror}
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\end{equation}
from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$. We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination. We constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21. We present results for bias in terminal year estimates of SSB and recruitment, Beverton-Holt stock recruit parameters ($a$ and $b$), and median natural mortality rate. Results for terminal year fishing mortality were strongly negatively correlated with those for SSB and are provided in the Supplementary Materials.

@@

We calculated Mohn's $\rho$ for SSB, fishing mortality (averaged over all age classes), and recruitment for each EM [@mohn99]. We fit $P =7$ peels to each simulated data set and calculated Mohn's $\rho$ for a given attribute $\theta$ as
\begin{equation}\label{mohns_rho}
\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
\end{equation}
where $\widehat \theta_{i,j}$ is the estimate for attribute $\theta$ in year $i$ from a model fit using data up to year $j$. For each EM we calculated median and 95\% confidence intervals using the same methods as that for relative bias.

@@

The measures of central tendency and variability of of observed values for specific OM and EM attributes (e.g., low or high observation error) that we described above can indicate scenarios that provide better or poorer reliability. However, the OM and EM attributes that we investigated are numerous, so we used two methods to summarize the most important factors for differences in results.  The first method was fitting regression models with the response being each of the measures of reliability described above and predictor variables were defined based on OM and EM characteristics [e.g., @mackinnonetal95; @wangetal17; @harwelletal18]. For the binary results for convergence and AIC-based selection of a stock-recruit relationship, we performed logistic regressions. For results for AIC-based selection of EM process error type (multiple categories) we performed multinomial regressions. For other measures of reliability we fit standard linear regression models. Because relative errors (Eq. \ref{relerror}) and Mohn's $\rho$ for the various parameters are bounded below at -1, we used a transformation of these values
\begin{equation}\label{bias_regression_response}
y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
\end{equation}
where $f$ is either the relative error (Eq. \ref{relerror}) or Mohn's $\rho$ (Eq. \ref{mohns_rho}) for simulation $i$, so that values are unbounded. For relative errors, $y_i$ is the log-scale error. We omitted simulation estimates equal to zero (RE = -1). For all regressions we fit separate models with individual factors included, with all factors combined, with including all second order interactions, and including all third order interactions. For the multinomial regression, we used the \verb|vglm| function from the VGAM package [@yee08;@yee15]. We tabulated percent reduction in residual deviance for each of regression fits. We did not perform formal statistical analyses of effects of OM and EM attributes on results (e.g., ANOVA) because of the lack of independence of the "observations" that results from fitting multiple EMs to each simulated data set.

The second method involved fitting classification and regression trees [@breimanetal84] to show how the OM and EM attributes, and their interactions, split the values for each measure of reliability [e.g., @gonzalezetal18;@collieretal22]. We used classification trees for categorical measures (convergence and AIC) and regression trees for the other measures with continuous scales (relative error and Mohn's $\rho$). The response variables were the same as the regressions for the deviance reduction analyses. We used the \verb|rpart| function in the rpart package (https://cran.r-project.org/package=rpart) to fit trees and, for clarity we pruned them to show the primary branches.

@@

### Deviance reduction and classification trees {-}

EM process error assumption was the single attribute that resulted in the largest percent reduction in deviance for all OM process error types other than R+S OMs where the EM M assumption explained the most residual deviance (Table \ref{convergence_PRD_table}). However, also including interactions of OM and EM factors also provided large reductions in residual deviance, suggesting successful convergence depended on a combination OM and EM attributes.  Classification trees for each OM process error type, all had the primary branch defined using the same attribute that provided the largest reduction in deviance (Figure \ref{conv_class}).

Generally, the factors defining primary splits for the trees corresponded to those with the largest reductions in deviance fitting the regression models.

### Full results {-}

For many R and R+S OMs, convergence rate declined when either the median natural mortality rate or the Beverton-Holt SRR was estimated even when the process error assumptions of the EMs and OMs matched (Figure \ref{hessian_SE_convergence}, A). When there was high observation error and constant fishing pressure ($F = F_{\text{MSY}}$ for all 40 years), convergence was poor for of all EM process error configurations other than R EMs when fitted to R OMs ($\sigma_{2+} = 0$) regardless of whether median natural mortality and SRRs were estimated. Convergence of R EMs was high for all R and R+S OMs except when there was high observation error and constant fishing pressure, and when median natural mortality and SRRs were estimated. R+S EMs fit to R OMs exhibited poor convergence regardless of whether natural mortality or a SRR was estimated. R+S EMs fit to R+S OMs had highest convergence rates when there was contrast in fishing pressure and low observation error. Convergence rates were high for all EMs when fit to data from R+S OMs with lower observation error except those where median natural mortality and/or SRRs were estimated.

Convergence of all EMs fitted to R+M OMs was highest when the OMs had higher natural mortality process error variability, low observation error, and contrast in fishing pressure  (Figure \ref{hessian_SE_convergence}, B). R+M EMs that estimated autocorrelation of process errors had poor convergence for R+M OMs when there was low natural mortality process error variability regardless of autocorrelation of the simulated process errors. R+S EMs fitted to data generated from R+M OMs always converged poorly whether or not median natural mortality and the Beverton-Holt SRR were estimated.

The R+S EMs, in particular, had poor convergence when fit to data generated from R+Sel OMs with lower selectivity process error variability or higher observation error (Figure \ref{hessian_SE_convergence}, C). R+Sel EMs generally converged better than other EMs for R+Sel OMs with higher process error variability, lower observation error, and contrast in fishing pressure regardless of whether median natural mortality or a SRR was estimated.

For R+q OMs, convergence of R+q EMs was generally better than that of other EMs when there was contrast in fishing pressure (Figure (\ref{hessian_SE_convergence}, D).  Convergence of R+S EMs was generally worse than that of all other EMs across all OMs whether or not median natural mortality or a SRR was estimated. Again, convergence probability generally declined for all EMs when median natural mortality or a SRR was estimated.

We found a wide range of maximum absolute values of gradients for models that converged (Figure \ref{hess_grad}). The largest value observed for a given EM and OM combination was typically $<10^{-3}$, but many converged models had values greater than 1. For many OMs, EMs that assumed the correct process error type and did not estimate median natural mortality or the Beverton-Holt SRR produced the lowest gradient values.

## AIC performance for process error structure {-}

### Deviance reduction and classification trees {-}

### Full results {-}

Marginal AIC accurately determined the correct process error assumptions in EMs when data were generated from R and R+S OMs, regardless of whether median natural mortality or a SRR was estimated (Figure \ref{pe_aic}, A). Attempting to estimate median natural mortality or a SRR separately had a negligible effect on the accuracy of determining the correct process error assumption. When both were estimated, there was a noticeable reduction in accuracy when OMs had a constant fishing pressure, low observation error, and larger variability in recruitment process errors.

For R+M OMs, marginal AIC only accurately determined the correct process error model and correlation structure when observation error was low and variability in natural mortality process errors was high (Figures \ref{pe_aic}, B). Of these OMs, estimating the median natural mortality rate only reduced the accuracy of AIC when natural mortality process errors were independent and fishing pressure was constant. For OMs with poor model selection accuracy, AIC most frequently selected EMs with process errors in catchability (R+q) or selectivity (R+Sel). Selection of R+S EMs was generally unlikely.

Marginal AIC most accurately determined the correct source of process error and correlation structure for R+Sel OMs with low observation error (Figures \ref{pe_aic}, C). When there was low variability in selectivity process errors and high observation error, R+q or R+S EMs were more likely to have the best AIC. Whether median natural morality or SRRs were estimated appeared to have little effect on the performance of AIC.

Marginal AIC most accurately determined the correct source of process error and correlation structure for R+q OMs with high variability in catchability process errors (Figures \ref{pe_aic},D). The R+q OMs with low variability in catchability process errors and high observation error had the least model selection accuracy. However, for these OMs, the marginal AIC accurately determined the correct source of process error (but not correlation structure) except when OMs assumed a constant fishing pressure and EMs estimated both median natural morality and the SRR.

## AIC performance for the stock-recruit relationship {-}

### Deviance reduction and classification trees {-}

### Full results {-}

Our comparisons of model performance conditioned on assuming the true process error configuration is known (EM and OM process error types match) and we focus on results where the EMs assume median natural mortality is known because there was little difference in results when the EMs estimated this parameter. Broadly, we found generally poor accuracy of AIC in selecting models assuming a Beverton-Holt SRR over the null model without an SRR for all OMs. However, we also found increased accuracy of AIC in determining the Beverton-Holt SRR when the simulated population exhibited greater variation in spawning biomass for nearly every OM (Figure \ref{sr_aic}).

With R and R+S process error assumptions, probability of lowest AIC for the B-H SRR as a function of SSB variability were greatest for OMs with contrast in fishing pressure and lower process variability in recruitment (Figure \ref{sr_aic}, A). The largest variation in SSB occurred in OMs with larger recruitment variability ($\sigma_R = 1.5$; Figure \ref{sr_aic}, A, right column group), but the same high AIC accuracy was achieved for OMs with lower recruitment variability at lower levels of SSB variation. The level of observation error had little effect on AIC accuracy.

For R+M OMs, probability of lowest AIC for the Beverton-Holt SRR increased steeply with variation in SSB whether it was induced by contrast in fishing or variation in natural mortality process error. (Figure \ref{sr_aic}, B). There was little difference in AIC accuracy whether the natural mortality process errors were correleted and, similar to R+S OMs, there was also little effect due to level of observation error.

For R+Sel OMs, contrast in fishing pressure over time was the primary source of variation in SSB and these are the OMs where AIC accuracy for the Beverton-Holt SRR was greatest (Figure \ref{sr_aic}, C). There was little effect of variability or correlation of selectivity process errors or the level of observation error on AIC accuracy.

Like the R+Sel OMs, the greatest accuracy for AIC in selecting the Beverton-Holt SRR occurred for R+q OMs where there was contrast in fishing pressure over time which is also where there was the greatest variation in SSB (Figure \ref{sr_aic}, D). There was also little effect of variability or correlation of catchability process errors or the level of observation error on AIC accuracy.

@@

### Deviance reduction and regression trees {-}

Across all OM process error types, fishing history was the most important single factor in reducing residual deviation in log-scale errors for the Beverton-Holt density-independent parameter $a$, and also for the density-independent parameter $b$ for R+M, R+Sel, and R+q OMs (Table \ref{bias_SR_pars_PRD_table}). For R and R+S OMs the EM process error assumption was equally important for estimating $b$, but the the reductions were small (mostly less than 4%) for any factor and any OM process error type.  There was also no substantial reduction in deviance with second or third order interactions of factors for either parameter in any OM process error types.

The regression trees indicate that having contrast in fishing pressure reduced the median error for both SR parameters for all OM types

We observed different effects of factors on deviance reduction for errors in estimation of the two Beverton-Holt stock-recruit parameters. When analyses were For bias of the density-dependent parameter $b$ as measured by the log absolute value of log-scale errors, deviance was not reduced by any of the individual OM or EM factors except for variation in recruitment $\sigma_R$ for R and R+S OMs. However, second and(or) third order interactions of the OM and EM factors also provided large reductions in deviance for all OM process error types which indicated that some combinations of factors could reduce bias for the $b$ parameter. Despite the explanation variation in the $b$ estimation erorrs, bias was still significant regardless of the combination of OM and EM factors (Figures regtrees and full results).

Median natural mortality rate showed reductions in deviance for R, R+S, and R+M OMs with second order interactions and R+M also showed reductions with third order interactions.

### Full results {-}

#### Spawning stock biomass {-}

For R OMs ($\sigma_{2+} = 0$), there was no indication of bias (95\% confidence intervals included 0) in terminal year SSB for any of the estimating models regardless of process error assumptions, except when no SR assumption was made, recruitment variability was low, and there was contrast in fishing mortality and high observation error (Figure \ref{SSB_rel_error}, A). However, errors in terminal SSB estimates were highly variable when median natural mortality was estimated and there was constant fishing pressure and high observation error (Figure \ref{SSB_rel_error}, A, second row).

For R+S OMs, the EMs with matching process error assumptions generally produced unbiased estimation of terminal SSB except when median natural mortality was estimated and there was high observation error. In R+S OMs with low observation error, EMs with incorrect process error assumptions typically provided biased estimation of terminal year SSB. Estimating the Beverton-Holt SRR had little discernible effect on bias of terminal year SSB estimation whereas estimating median M tended to produce more variability in errors in terminal SSB estimation similar to R OMs.

For R+ M OMs with low variability in natural mortality process errors, low observation error and contrast in fishing motality over time all EMs produced low variability in SSB estimation error that indicated unbiasedness (Figure \ref{SSB_rel_error}, B, third row). However, larger variability in natural mortality process errors increased bias of EMs without the correct process error type. Estimating median natural mortality increased variability of SSB estimation error particularly for OMs with high observation error and constant fishing pressure over time.  It also increased bias in SSB estimation for many R+M OMs. Like R and R+S OMs, estimating a SRR had little discernible effect on SSB bias.

For R+Sel OMs, there was no evidence of bias for any EMs when variability in selectivity process error and observation error was low, and with contrast in fishing mortality (Figure \ref{SSB_rel_error}, C). The largest bias occurred for any EMs that estimated median natural mortality when the OMs had high observation error, constant fishing pressure, and greater variability in selectivity process errors ($\sigma_{\text{Sel}} = 0.5$) or low selectivity process errors ($\sigma_{\text{Sel}} = 0.1$) and low observation error. However, there was no evidence of SSB bias for correctly specified R+Sel EMs when observation error was low and variation in selectivity process errors was larger, whether median natural mortality was estimated or not (Figure \ref{SSB_rel_error}, C, third row). We only observed an effect of estimating the Beverton-Holt SRR for R+Sel OMs that had high observation error and contrast in fishing pressure where estimating the SRR produced less biased SSB estimation for many EMs (Figure \ref{SSB_rel_error}, C, top row).

All EMs fit to data from R+q OMs with low observation error and  contrast in fishing pressure exhibited little evidence of bias in terminal SSB estimation except for R+M EMs when there was no AR1 correlation in catchability process errors (Figure \ref{SSB_rel_error}, D). Many EMs also performed well in R+q OMs with low observation error, but no contrast in fishing pressure. For R+q OMs with high observation error and contrast in fishing pressure, EMs that estimated the Beverton-Holt SRR exhibited less SSB bias than those that did not. Estimating median natural mortality in the EMs only resulted in much more variable SSB estimation errors when there was no contrast in fishing pressure (Figure \ref{SSB_rel_error}, D, first and third rows).

For all OM process error types, relative errors in terminal year recruitment were generally more variable than SSB, but effects of R and R+S OM and EM attributes on bias (i.e, negative or positive or none) were similar (Figure \ref{R_rel_error}, A). Furthermore, for EM configurations where bias in terminal SSB was evident, median relative errors in recruitment often indicated stronger bias in recruitment of the same sign.

### Beverton-Holt parameters{-}

Across all OMs, there was generally less bias and(or) lower variability in estimation of the Beverton Holt $a$ parameter than the $b$ parameter. In R and R+S OMs, EMs with the correct assumptions about process errors provided the least biased estimation of Beverton-Holt SRR parameters when there was a change in fishing pressure over time and lower variability of recruitment process errors, but there was little effect of estimating median natural mortality and a small increase in bias for those OMs that had high observation error (Figure \ref{SR_rel_error}, A). For other R and R+S OMs, estimating natural mortality often resulted in less biased estimation of SRR parameters. There was generally large variability in relative errors of the SRR parameter estimates, but the lowest variability occurred with low variability in recruitment and little or no variability in survival process errors ($\sigma_{2+} \in \{0,0.25\}$), and contrast in fishing pressure.

In R+M OMs, the most accurate estimation of SRR parameters for all EM process error assumptions occurred when there was a change in fishing pressure, greater variability in natural mortality process errors, and lower observation error (Figure \ref{SR_rel_error}, B). Relative to the R, and R+S OMs, there was even less effect of estimating median natural mortality on estimation bias for the SRR parameters.

Bias for SRR parameters was large and variability in relative errors was greatest for most EMs fit to R+Sel OMs with constant fishing pressure (Figure \ref{SR_rel_error}, C). Less bias in parameter estimation occurred for OMs with a change in fishing pressure and the best accuracy occurred for those OMs that had low observation error and more variable and uncorrelated selectivity process errors, and when the EMs had with the correct process error assumption. There was little effect of estimating natural mortality on relative errors for SRR parameters.

>>>>

State-space models are increasingly used for stock assessment, and evaluations of statistical inferences made from them is needed. We simulated 72 operating models that varied fishing pressure and observation error across process errors in recruitment, apparent survival, selectivity, catchability, and/or natural mortality. We fit estimating models with different assumptions on the process error source and whether natural mortality and stock-recruit relationship were estimated. We measured reliability of estimating models by probability of convergence, accuracy of AIC for correct process error configuration and stock-recruit function, estimation bias, and magnitude of retrospective patterns. Across all measures reliability was generally better with lower observation error, contrast in fishing pressure over time, and median natural mortality rate known, but using marginal AIC to distinguishing the stock recruit relationship further required large contrast in spawning biomass and low residual variation in recruitment. Marginal AIC accurately distinguished process errors on recruitment, apparent survival, and selectivity.  These results improve our understanding of when results from state space assessment models, the next-generation of fisheries stock assessment, can be expected to be reliable.

@@

Subsets of 20 EMs in Table \ref{em_table} were fit to simulate data sets from each of the OM process error categories. For R and R+S OMs, fitted EMs had matching process error assumptions as well as R+Sel, R+M, and R+q assumptions without autocorrelation. Similarly, For other OM process error categories, we fit EMs with matching process error assumptions as well as other process error types without autocorrelation. The maturity at age, weight at age for catch and spawning stock biomass (SSB), and observation error variance of aggregate catch and indices were all assumed known at the true values. However, the variance parameters for the logistic-normal distributions for age composition observations were estimated in the EMs. As such, EMs would either be configured completely correctly for the OM, or there could be mis-specification in assumptions of process error autocorrelation, the type of process error, or the SRR (Beverton-Holt or none).

@@

The first measure of reliability we investigated was frequency of convergence when fitting each EM to the simulated data sets. There are various ways to assess convergence of the fit [e.g., @carvalhoetal21; @kapuretal24], but given the importance of estimates of uncertainty when using assessment models in management, we estimated probability of convergence as measured by occurrence of a positive-definite hessian matrix at the optimized negative log-likelihood that could be inverted (i.e., providing hessian-based standard error estimates). We also provide results in the Supplementary Materials for convergence defined by the maximum absolute gradient < $1^{-6}$ and the maximum of the absolute gradient values for all fits of a given EM to all simulated data sets from a given OM that produced hessian-based standard errors for all estimated fixed effects. This provides an indication of how poor the calculated gradients can be, but still presumably converged adequately enough for parameter inferences.

@@

We investigated the reliability of AIC-based model selection for two purposes. First, we analyzed selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given OM simulated data set, we compared AIC for EMs with different process error assumption conditional on whether median natural mortality rate and the Beverton-Holt SRR were estimated. We tabulated the models providing the lowest AIC across simulated data sets. Second, we analyzed AIC-based selection between EMs with and without the Beverton-Holt SRR assumed. Contrast in fishing pressure and time series with recruitment at low stock size has been shown to improve estimation of SRR parameters [@magnussonhilborn07; @connetal10]. Our preliminary inspections indicated generally poor performance of AIC in determining the Beverton-Holt model for a given set of OM factors (including contrast in fishing pressure). Therefore, we also considered the effect of of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) on model selection since simulations with realized SSB producing low and high recruitment would have larger variation in realized SSB.

All model selection results condition only on completion of the optimization process without failure for all of the compared EMs. We did not condition on convergence as defined above because optimization could correctly determine an inappropriate process error assumption by estimating variance parameters at the lower bound of zero. Such an optimization could indicate poor convergence but the likelihood would be equivalent to that without the mis-specified random effects and the AIC would be appropriately higher because more (variance) parameters were estimated. All other measures of reliability described below (bias and Mohn's $\rho$) use these same criteria for inclusion of EM fits in the summarized results.

@@

We also investigated bias in estimation of various model attributes as a measure of reliability. For a given model attribute we calculated the relative error 
\begin{equation}\label{relerror}
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\end{equation}
from fitting a given estimating model to simulated data set $i$ configured for a given OM where $\widehat \theta_i$ and $\theta_i$ are the estimated and true values for simulation $i$. We analyzed simulation results for estimates of terminal year estimates of SSB and recruitment, Beverton-Holt SRR parameters ($a$ and $b$), and median natural mortality rate.

@@

Finally, we investigated presence of retrospective patterns in fitted models as a measure of reliability. We calculated Mohn's $\rho$ for SSB, fishing mortality (averaged over all age classes), and recruitment for each EM fit to each OM simulated data set [@mohn99]. We fit $P = 7$ peels to each simulated data set and calculated Mohn's $\rho$ for a given attribute $\theta$ as
\begin{equation}\label{mohns_rho}
\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
\end{equation}
where $\widehat \theta_{i,j}$ is the estimate for attribute $\theta$ in year $i$ from a model fit using data up to year $j$.

@@

The measures of central tendency and variability of observed values for specific OM and EM attributes (e.g., low or high observation error) that we described above can indicate scenarios that provide better or poorer reliability. However, the OM and EM attributes that we investigated are numerous, so we used two methods to summarize the most important factors for differences in results.  The first method was fitting regression models with the response being each of the measures of reliability described above and predictor variables were defined based on OM and EM characteristics [e.g., @mackinnonetal95; @wangetal17; @harwelletal18]. For the binary indicators of convergence and AIC-based selection of a SRR, we performed logistic regressions. For indicators of AIC-based selection of EM process error type (multiple categories) we performed multinomial regressions. For other measures of reliability we fit linear regression models to transformed responses. Because relative errors (Eq. \ref{relerror}) and Mohn's $\rho$ for the various parameters are bounded below at -1, we used a transformation of these values
\begin{equation}\label{bias_regression_response}
y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
\end{equation}
where $f$ is either the relative error (Eq. \ref{relerror}) or Mohn's $\rho$ (Eq. \ref{mohns_rho}) for simulation $i$, so that values are unbounded. For relative errors, $y_i$ is the log-scale error. We omitted simulation estimates equal to zero (RE = -1). For all regressions we fit separate models with individual factors included, with all factors combined, with including all second order interactions, and including all third order interactions. For the multinomial regression, we used the \verb|vglm| function from the VGAM package [@yee08;@yee15]. We tabulated percent reduction in residual deviance for each of regression fits. We did not perform formal statistical analyses of effects of OM and EM attributes on results (e.g., ANOVA) because of the lack of independence of the "observations" that results from fitting multiple EMs to each simulated data set.

The second method involved fitting classification and regression trees [@breimanetal84] to show how the OM and EM attributes, and their interactions, split the values for each measure of reliability [e.g., @gonzalezetal18;@collieretal22]. We used classification trees for categorical measures (convergence and AIC) and regression trees for the other measures with continuous scales (relative error and Mohn's $\rho$). The response variables were the same as the regressions for the deviance reduction analyses. We used the \verb|rpart| function in the rpart package (https://cran.r-project.org/package=rpart) to fit trees. Full trees were determined using default settings except that we increased the number of cross-validations to 100. For clarity, we pruned the full trees to show just the primary branches.

We also provide detailed results for all measures of reliability at each combination of OM and EM attributes in the Supplementary Materials. For confidence intervals of probability of convergence, we used the Clopper-Pearson exact method [@clopperpearson34;@thulin14]. For AIC selection of process error configuration we provide estimates of the proportions of simulations where each EM type was selected. For AIC selection of the SRR, we provide predicted probabilities from logistic regressions as a function of SSB variability for each OM and EM type. We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination. We constructed 95\% confidence intervals for the median relative bias, and Mohn's $\rho$ using the binomial distribution approach as in @millerhyun18 and @stockmiller21.  For each EM we calculated median and 95\% confidence intervals using the same methods as that for relative bias.

@@

For probability of convergence, EM process error assumption was the single attribute that resulted in the largest percent reduction in deviance (14-28\%) for all OM process error types other than R+S OMs where the EM M assumption explained the most residual deviance (>11\%; Table \ref{convergence_PRD_table}). However, including interactions of OM and EM factors also provided large reductions in residual deviance (35-47\%), suggesting successful convergence depended on a combination OM and EM attributes.

Classification trees for each OM process error type, all had the primary branch defined using the same attribute that provided the largest reduction in deviance (Figure \ref{conv_class}). EMs that assumed R+S process errors converged poorly for all OMs that were simulated with the alternative process error assumptions (R, R+M, R+Sel, an R+q OMs). For all trees, branches based on the OM fishing mortality history showed better convergence when the OM included a change in fishing pressure. Branches based on whether the Beverton-Holt SRR was assumed or not, showed better convergence when it was not estimated and branches based on whether the EM estimated the median natural mortality rate showed better convergence when it was treated as known. For certain R+M and R+Sel OMs, better convergence was also observed when there was lower observation uncertainty.

When convergence is defined by a gradient threshold, the primary factor explaining deviance reduction is the same for all OM process error types, but there are some differences in deviance reduction for secondary factors (Table \ref{convergence_gradient_PRD_table}), and probability of convergence, overall, was lower (Figure \ref{conv_gradient_class}). We found a wide range of maximum absolute values of gradients for models that had invertible Hessians (Figure \ref{hess_grad}). The largest value observed for a given EM and OM combination was typically $<10^{-3}$, but many converged models had values greater than 1. For many OMs, EMs that assumed the correct process error type and did not estimate median natural mortality or the Beverton-Holt SRR produced the lowest gradient values.

## AIC performance {-}

### Process error structure {-}

For AIC selection of the correct process error configuration, the magnitude of observation and process error were the attributes that resulted in the largest percent reductions in deviance across OM process error types other than R OMs (Table \ref{AIC_PE_PRD_table}). Both variance of apparent survival random effects ($\sigma_{2+}$) and degree of observation error explained the largest reductions for R+S (17-22\%) and R+Sel (8-26\%) OMs, whereas variance of process errors provided the largest reductions in R+M (>9\%) and R+q (>13\%) OMs. Comparatively, none of the OM or EM attributes explained particularly large reductions in deviance for R OMs, but fishing history, whether a SRR was estimated, and whether median natural mortality was known or estimated provided similar and the largest reductions (approximately 5-6\%). Inclusion of second and third order interactions, did not provide large reductions in deviance for any of the OM process error types.

For all OM process error types other than R OMs, the attributes defining the primary branches of classification trees matched those that provided the largest reductions in deviance (Figure \ref{AIC_PE_class}). Across all OMs, AIC was more accurate for the process error type when process error variability was greater and when observation error was lower. No branches were estimated for classification trees fit to the R OMs, likely because accuracy was high across all simulations (0.94), although inspection of the fine-scale results shows there is some degradation in AIC selection when a SRR and median natural mortality rate are estimated for R OMs with constant fishing pressure and high observation error (Figure \ref{pe_aic}, top left).

### Stock-recruit relationship {-}

Logistic regressions for AIC selection of the Beverton-Holt SRR, showed OM fishing history and variation in SSB ($\log \text{SD}_{\text{SSB}}$) provided substantial reductions in deviance for R+M (>13\%), R+Sel (>26\%), and R+q (>24\%) OMs (Table \ref{AIC_SRR_PRD_table}). For R OMs, fishing history provided the largest reduction in deviance (>9\%) whereas none of the attributes individually provided large reductions in deviance for R+S OMs (all <5\%). However, inclusion of all attributes provided larger reductions in deviance than the sum of individual contributions for both R (>30\%) and R+S (>18\%) OMs. For all OM process error types, inclusion of interaction terms provided relatively little reduction in residual deviance.

Attributes defining the primary branches of classification trees for AIC selection of the SRR assumption were the same as those explaining the largest reductions in deviance for the logistic regression models (Figure \ref{AIC_SRR_class}). All branches based on variation in SSB showed better accuracy with larger variability in SSB and all branches based on fishing history showed better accuracy when there was contrast in fishing pressure. Branched based on OM observation error or recruitment variability showed better accuracy when they were lower. For R OMs, a combination of lower recruitment variability, contrast in fishing pressure, and higher SSB variability produced AIC accuracy over 0.8. For R+S OMs, lower recruitment variability and observation error and higher SSB variability produced AIC accuracy of 0.79. For R+M, R+Sel, and R+q OM,s accuracy of 0.87 to 0.94 was observed just with higher SSB variability.

@@

### Terminal year spawning stock biomass, fishing mortality, and recruitment {-}

Regression models for log-scale errors in SSB that included the various OM and EM factors showed little reduction in deviance (<5\%) for any of the factors across all OM process error types (Table \ref{bias_SSB_PRD_table}). The attributes producing the largest reductions were the EM assumption for median natural mortality (known or estimated) for R, R+M, R+Sel, and R+q OMs (1-3\%), EM process error type for R+S OMs (4\%) and fishing history for all OM types (1-5\%). Including second interactions provided largest reductions in residual deviance (10- 26\%). Including third order interactions also provided large further reductions for R, R+S, and R+q OMs (between 5 and 11\%). In all regression trees, branches based on fishing history and level of observation error generally showed lower bias in SSB with contrast in fishing and lower observation error (Figure \ref{SSB_bias_regtree}). For branches based on treatment of median natural mortality rate, bias was generally lower when it was known rather than estimated. For some R+Sel and R+q OMs, less bias in SSB was shown when the EM process error configuration was correct.

Results for bias in fishing mortality and recruitment generally matched those for SSB, except that directions of bias for fishing mortality were opposite to those for SSB and recruitment. Effects of individual OM and EM factors on regression models were similarly small as measured by reduction in deviance (Tables \ref{bias_F_PRD_table} and \ref{bias_R_PRD_table}). Primary branches of regression trees were in most cases identical to those for SSB  (Figures \ref{F_bias_regtree} and \ref{R_bias_regtree}).

### Stock-recruit parameters {-}

Regression models for transformed relative errors of estimates of both the Beverton-Holt $a$ and $b$ parameters showed none of the factors explained large percent reductions in deviance (Table \ref{bias_SR_pars_PRD_table}). The OM fishing history provided the largest reduction for most OM process error types for both parameters, but percent reductions were less than 5.6\% except for R+Sel OMs where the percent reductions were 11.37\% and 7.97\% for the $a$ and $b$ parameters, respectively and for just the $b$ parameter for R+q OMs (10\%). The EM process error assumption provided similar reductions in deviance for both parameters for R OMs. Including interactions also did not produce important reductions in deviance (increase of .

For regression trees of log-scale error in Beverton-Holt $a$ and $b$ parameters, lower bias was indicated with contrast in OM fishing pressure for all branches in trees for each OM process error type (Figures \ref{SR_a_bias_regtree} and \ref{SR_b_bias_regtree}). For all branches based on recruitment variability in trees for R and R+S OMs, lower bias in both $a$ and $b$ was observed with less recruitment variability. For R OMs with contrast in fishing pressure and greater recruitment variability EMs that assumed the incorrect R+M process errors produced lower bias in both $a$ and $b$ than other process error assumptions.  Across all OMs, there was generally less bias and(or) lower variability in estimation of the $a$ parameter than the $b$ parameter (Figure \ref{SR_rel_error}).
